---
title: "PA1_template"
author: "Bishnu Rawal"
date: "April 26, 2016"
output: html_document
---
#Basic settings, loading data, calculating mean and median of steps each day
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load the data from the source file
```{r }
# Read activity.csv file into "activity" variable
activity <- read.csv(file="activity.csv")
# Convert date field from factor to date
activity$date <- as.Date(activity$date)
```

###Process data to make it ready for plotting and analysis
```{r}
# Load reshape2 library to get melt & dcast functions
library(reshape2)

# Melt data frame to prep for casting by date -- by setting the id variable to date and the measure variable to steps, we get a table with multiple values for steps taken within each day
actMeltDate <- melt(activity, id.vars="date", measure.vars="steps", na.rm=FALSE)

# Cast data frame to see steps per day -- this sums the steps by date to give us a table of 3 columns x 61 rows
actCastDate <- dcast(actMeltDate, date ~ variable, sum)

```

###Plot histogram of data
```{r actCastDate, echo=FALSE}
# Plot histogram with frequency of steps by day and add a red line showing the mean value
plot(actCastDate$date, actCastDate$steps, type="h", main="Histogram of Daily Steps", xlab="Date", ylab="Steps per Day", col="blue", lwd=8)
abline(h=mean(actCastDate$steps, na.rm=TRUE), col="red", lwd=2)
```
###Calculate mean and median of steps by day
```{r}
# Calculate mean and median of daily steps
paste("Mean Steps per Day =", mean(actCastDate$steps, na.rm=TRUE))
paste("Median Steps per Day =", median(actCastDate$steps, na.rm=TRUE))
```
#Calculating and plotting the average daily activity pattern by interval

###Reprocess data to calculate by interval instead of day
```{r}
# Re-melt data frame to prep for casting by interval, including removing NA values so we can take the mean a little later
actMeltInt <- melt(activity, id.vars="interval", measure.vars="steps", na.rm=TRUE)

# Cast data frame to see mean steps per interval
actCastInt <- dcast(actMeltInt, interval ~ variable, mean)
```
###Create a time series plot of average steps by interval
```{r}
# Plot line chart with average frequency of steps by interval and add line with mean
plot(actCastInt$interval, actCastInt$steps, type="l", main="Frequency of Steps Taken at Each Interval", xlab="Interval ID", ylab="Steps", col="orange", lwd=2)
abline(h=mean(actCastInt$steps, na.rm=TRUE), col="red", lwd=2)

# Output interval that has max value along with the max value
paste("Interval with max value =", actCastInt$interval[which(actCastInt$steps == max(actCastInt$steps))])
paste("Maximum interval mean steps =", max(actCastInt$steps))
```

#Imputing missing values to replace NAs in data set and compare results

###Calculate total number of missing values in dataset
```{r}
sum(is.na(activity$steps))
```
###Document strategy of dealing with missing value
There are a large number of missing/NA values (2,304), I will replace NAs with the mean for the particular interval number. For example: if the average number of steps taken during interval x is y, I will replace each NA with the corresponding y value for that row). I will then recalculate the steps per day to see how much it differs from the original result (with NAs included), if at all.

###Create new data set with imputed NA values as stated in strategy
```{r}
# Data frame with mean steps per interval - just renaming to be more descriptive
stepsPerInt <- actCastInt

# Create data frame that we will remove NAs from
actNoNA <- activity

# Merge activity data set with stepsPerInt data set
actMerge = merge(actNoNA, stepsPerInt, by="interval", suffixes=c(".act", ".spi"))

# Get list of indexes where steps value = NA
naIndex = which(is.na(actNoNA$steps))

# Replace NA values with value from steps.spi
actNoNA[naIndex,"steps"] = actMerge[naIndex,"steps.spi"]
```
###Plot histogram and calculate mean and median of total steps/day with new (no NA) data set and compare with original.
```{r}
# Melt new data frame to prep for casting by date
actMeltDateNoNA <- melt(actNoNA, id.vars="date", measure.vars="steps", na.rm=FALSE)

# Cast data frame to see steps per day
actCastDateNoNA <- dcast(actMeltDateNoNA, date ~ variable, sum)

# Plot histogram with frequency of steps by day
plot(actCastDateNoNA$date, actCastDateNoNA$steps, type="h", main="Histogram of Daily Steps (Imputted NA Values)", xlab="Date", ylab="Steps", col="gray", lwd=8)
abline(h=mean(actCastDateNoNA$steps), col="red", lwd=2)
# Calculate mean and median of daily steps
paste("Mean daily steps =", mean(actCastDateNoNA$steps, na.rm=TRUE))
paste("Median daily steps =", median(actCastDateNoNA$steps, na.rm=TRUE))
```
#Check if differences exist in activity patterns between weekdays and weekends
###Create a factor variable that states whether each day is a weekday or weekend
```{r}
# For loop to create new column called "dayOfWeek" and insert whether each date corresponds to a weekday or weekend
for (i in 1:nrow(actNoNA)) {
    if (weekdays(actNoNA$date[i]) == "Saturday" | weekdays(actNoNA$date[i]) == "Sunday") {
        actNoNA$dayOfWeek[i] = "weekend"
    } else {
        actNoNA$dayOfWeek[i] = "weekday"
    }
}
```
###Subset, process, and plot two charts (one above the other) to compare weekday vs. weekend activity
```{r}
# To create a plot, we must first subset the data
actWeekday <- subset(actNoNA, dayOfWeek=="weekday")
actWeekend <- subset(actNoNA, dayOfWeek=="weekend")

# Next, we need to process the data for our needs
actMeltWeekday <- melt(actWeekday, id.vars="interval", measure.vars="steps")
actMeltWeekend <- melt(actWeekend, id.vars="interval", measure.vars="steps")
actCastWeekday <- dcast(actMeltWeekday, interval ~ variable, mean)
actCastWeekend <- dcast(actMeltWeekend, interval ~ variable, mean)

# Load plot packages necessary to continue
library(ggplot2)
library(gridExtra)
# Set plot area to two rows and one column, and then plot charts with mean line in each
plot1 <- qplot(actCastWeekday$interval, actCastWeekday$steps, geom="line", data=actCastWeekday, main="Steps by Interval - Weekday", xlab="Interval ID", ylab="Number of Steps")
plot2 <- qplot(actCastWeekend$interval, actCastWeekend$steps, geom="line", data=actCastWeekend, main="Steps by Interval - Weekend", xlab="Interval ID", ylab="Number of Steps")
grid.arrange(plot1, plot2, nrow=2)
```
